---
title: "Power analysis of scRNA-seq experiment"
author: ""
date: "`r Sys.Date()`"
output:
        html_document:
                toc: true
                toc_float: true
                collapsed: true
                toc_depth: 2
                theme: paper
---

```{r setup, cache=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Load libraries
library(Seurat)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(Matrix)
library(AnnotationHub)
library(ensembldb)
library(scales)
library(SingleCellExperiment)

# Set seed for reproducibility
set.seed(1454944673L)
opts_chunk[["set"]](
    audodep = TRUE,
    cache = FALSE,
    cache.lazy = FALSE,
    error = TRUE,
    echo = FALSE,
    fig.height = 10L,
    fig.retina = 2L,
    fig.width = 10L,
    message = FALSE,
    tidy = TRUE,
    warning = TRUE
)

```

# Power analysis

Power analyses were performed to determine the number of samples needed to identify differentially expressed genes between `samplegroup1` and `samplegroup2` given a false positive rate of 0.05 (alpha) and a false negative rate of 0.1 or 0.2 (80% or 90% power).

For single-cell analyses with multiple biological replicates, differential expression analysis can be performed by summing the single cells for each gene to the sample level for the cell type of interest [[1](https://goo.gl/hWpXQF)]. Therefore, to determine the number of required biological replicates needed to achieve a specific power to detect differentially expressed genes for a given effect size, the single cells are collapsed to the sample level to perform a pseudo-bulk power analysis.

To perform the power analysis, the power was calculated by calculating the average number of counts per gene in the cell type population of interest and exploring a range of effect sizes using the RNASeqPower package in R [[2](https://www.bioconductor.org/packages/release/bioc/vignettes/RNASeqPower/inst/doc/samplesize.pdf)]. With the pilot analysis containing a single patient sample, we cannot calculate the biological coefficient of variation (CV); however, within group CV has been found to be less than 0.4 for a range of human studies in bulk RNA-seq [[3](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3842884/)]. Since single-cell RNA-seq analyses have additional variation, we will use a stringent CV, set to 0.5. It is likely a conservative estimate of variability between patients, but we will be able to better estimate the the variability after sequencing additional patients.

Our cell population of most interest is `celltype`. We performed the power calculation for this population first. Below are the sample sizes needed for 80% and 90% power and a range of effect sizes given on the left:

```{r power_analysis, message=FALSE, warning=FALSE}
# Load library for power analysis
library(RNASeqPower)
library(basejump)

# Bring in clustering information
seurat_tsne <- readRDS("path/to/seurat.rds")

metadata <- seurat_tsne@meta.data
metadata$ident <- seurat_tsne@ident

# Subset cells to cluster of interest
# Example Monocyte cells

cellIDs_monocytes <- rownames(metadata[metadata$ident == "Monocytes", ])
monocyte_counts <- seurat_tsne@raw.data[, cellIDs_monocytes]
#all(cellIDs_monocytes %in% colnames(monocyte_counts))

# Determine average number of reads per gene for each sample
total_gene_reads <- rowSums(as.matrix(monocyte_counts))
avg_gene_reads <- geometricMean(total_gene_reads)

# Determine the biological coefficient of variation
## If only a single sample and can't estimate, use bcv of 0.5
bcv <- 0.5

## If more than a single sample:

library(edgeR)
edgeR_dgelist <- DGEList(counts=monocyte_counts, group=condition))
edgeR_dgelist <- calcNormFactors(edgeR_dgelist, method = "TMM")
edgeR_dgelist <- estimateCommonDisp(edgeR_dgelist)
bcv <- sqrt(edgeR_dgelist$common.dispersion) 

# Run the power analysis
power_table <- rnapower(depth=avg_gene_reads,
                        cv=bcv, 
                        effect=c(1.5, 1.75, 2),
                        alpha= .05, 
                        power=c(.8, .9))
knitr::kable(power_table)
```

To achieve 80% power for identifying of genes differentially expressed between `samplegroup1` and `samplegroup2` monocytes by at least 2-fold would require x replicates per sample group. If we wanted to achieve 90% power, then we would require x replicates. Similarly, if we preferred to identify DE genes with 1.75-fold differences in expression, then we would need at least x replicates for 80% power and x replicates for 90% power. 

Since monocytes are one of the larger cell populations we could explore with 9,674 cells, we could also test one of the smaller cell populations. 

```{r cell_pop}
cell_freqs <- data.frame(sort(table(metadata_all$ident), decreasing = TRUE))
colnames(cell_freqs) <- c("Cell type", "Number of cells")
knitr::kable(cell_freqs)
```

For the T cells with x cells, we can perform the same power analysis:

```{r power_analysis_Tcells, message=FALSE, warning=FALSE}
# Load library for power analysis
# Subset cells to cluster of interest
# Example T cells

cellIDs_Tcells <- rownames(metadata[metadata$ident == "T cells", ])
Tcells_counts <- seurat_tsne@raw.data[, cellIDs_Tcells]
#all(cellIDs_Tcells %in% colnames(Tcells_counts))

# Determine average number of reads per gene for each sample
total_gene_reads <- rowSums(as.matrix(Tcells_counts))
avg_gene_reads <- geometricMean(total_gene_reads)

# Determine the biological coefficient of variation
## Since only a single sample and can't estimate, use bcv of 0.5
bcv <- 0.5

# Run the power analysis
power_table <- rnapower(depth=avg_gene_reads,
                        cv=bcv, 
                        effect=c(1.5, 1.75, 2),
                        alpha= .05, 
                        power=c(.8, .9))
knitr::kable(power_table)
```

With the fewer number of cells, we would need more samples to detect differences in expression between the groups. For 80% power, we would need x replicates to detect an effect size of 2 and x replicates to detect an effect size of 1.75.

**References:**

1. H.M. Kang, M. Subramaniam, S. Targ, M. Nguyen, L. Maliskova, E. Wan, S. Wong, L. Byrnes, C. Lanata, R. Gate, et al. Multiplexing droplet-based single cell RNA-sequencing using natural genetic barcodes. Nature Biotechnology (2018), 36:89–94. Doi: 10.1038/nbt.4042  

2. Terry Therneau, Steven Hart and Jean-Pierre Kocher (2019). Calculating
  samplesSize estimates for RNA Seq studies. R package version 1.22.1.
  
3. Hart SN, Therneau TM, Zhang Y, Poland GA, Kocher JP. Calculating sample size estimates for RNA sequencing data. J Comput Biol. (2013), 12:970-8. Doi: 10.1089/cmb.2012.0283. 
